---
title: "Introduction of geneHapR"
author: "Zhang RenLiang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction of geneHapR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`geneHapR` is an R package that designed for gene haplotype statistics, phenotype association and visualization.

## Preliminaries

Input files frequently present challenges to analysis.
A common problem I encounter is that chromosome names and gene IDs are not standardized among **VCF**, **FASTA** and **GFF** files.
That means additional work for users with or without R.
I suggest that reading these files into R, synchronizing the names in R and then proceeding with downstream analyses.

Large **VCF** or **GFF** file is another important consideration when using `geneHapR.` 
Usually, the annotation file (GFF) and variants call format (VCF) files are commonly at chromosome or genome level. 
Thus those files may be huge. 
However, a gene size usually range from hundreds to thousands bp.
It means the dataset for a gene haplotype analysis is very small portion inside the huge file. Thus I have developed `filterLargeVCF()` (see **VCF filtration**) in case the user known a little or nothing about linux. However, I suggest to extract the target ranges from the whole VCF or GFF file with in linux software before import them into R if the user is familiar with linux. It will save much time and brainwork if the user have hundred genes or ranges need to be extract from the original file.

## Installation

`geneHapR` could be installed from **gitee** (preferred) or **github**.
More information about `geneHapR` could be found at [https://gitee.com/zhangrenl/genehapr](https://gitee.com/zhangrenl/genehapr)
 
```{r eval=FALSE}
devtools::install_git("https://gitee.com/zhangrenl/geneHapR")
devtools::install_github("zhangrenl/geneHapR")
```

`geneHapR` is schemed to submit to CRAN. 
`geneHapR` has not published yet, if you use `geneHapR` in your study, please contact [Zhang RenLiang (email: zhang_renliang@163.com)](mailto: zhang_renliang@163.com) or [Jia GuanQing (jiaguanqing@caas.cn)](mailto: jiaguanqing@caas.cn)

## Data input

The first step is library the `geneHapR` packages.

```{r startup, echo=TRUE, message=FALSE, warning=FALSE}
library(geneHapR)
```

```{r setups, include=FALSE}
data("geneHapR_test")
```

There are two options to conduct a gene haplotype analysis starts from a VCF file or DNA sequcences file. Thus a **VCF** file or **DNA sequences** file is necessary. However, the *GFF*, *phenos* and *accession groups* are strongly recommend for visualization and phenotype associations.

The import functions takes file name/path as input. `import_vcf()` could import VCF file with surfix of ".vcf" and ".vcf.gz". `import_gff()` import file format default as "GFF" and `import_seqs()` file format default as "fasta".

### Import VCF

```{r import vcf, eval=FALSE, include=TRUE}
# import vcf file
vcf <- import_vcf("your_vcf_file_path.vcf")

# import gziped vcf file
vcf <- import_vcf("your_vcf_file_path.vcf.gz")
```

### Import GFF

```{r import gff, eval=FALSE, include=TRUE}
# import GFFs
gff <- import_gff("your_gff_file_path.gff", format = "GFF")
```

### Import DNA sequences

```{r import DNA seqs, eval=FALSE, include=TRUE}
# import DNA sequences in fasta format
seqs <- import_seqs("your_DNA_seq_file_path.fa", format = "fasta")
```

## Import phenotype and accession group information

```{r import phenotype and accession groups, eval=FALSE, include=TRUE}
# import phynotype data
pheno <- import_pheno("your_pheno_file_path.txt")
pheno
```

```{r echo=FALSE}
head(pheno)
```

```{r import pheno, eval=FALSE}

# import accession group information
accGroup <- import_pheno("accession_group_file_path.txt")
```

```{r echo=FALSE}
head(accGroup)
```

Be aware that the phenotype and accession group are effectively tables. There are more than one ways to import a table format file with `R`. 

Just **Note** that: 
a. the accession/individual names located in first column; 
b. the first row contents phenotype/accession_group names; 
c. `NA` is allowed, it's not a wise option to replace `NA` by `0`.

eg.

```{r import pheno and accessions_groups, eval=FALSE, include=TRUE}
# import pheno from space ' ' delimed table
pheno <- read.table("your_pheno_file_path.csv", header = TRUE, row.names = 1, comment.char = "#")

# import pheno from ',' delimed table
pheno <- read.csv("your_pheno_file_path.csv", header = TRUE, comment.char = "#")
```

## Data manipulations

There is a little work need to be done before haplotype calculations: (1) VCF filtration and (2) DNA sequences alignment.

### VCF filtration

There are three modes to filter a `vcfR` object after import VCF into R: a. by position; b. by annotation; c. by both of them.

```{r VCF filtration, eval=FALSE, include=TRUE}
# filter VCF by position
vcf_f1 <- filter_vcf(vcf, mode = "POS",
                     Chr = "scaffold_1",
                     start = 4300, end = 5890)

# filter VCF by annotation
vcf_f2 <- filter_vcf(vcf, mode = "type",
                     gff = gff,
                     type = "CDS")

# filter VCF by position and annotation
vcf_f3 <- filter_vcf(vcf, mode = "both",
                     Chr = "scaffold_1",
                     start = 4300, end = 5890,
                     gff = gff,
                     type = "CDS")
```

It's a time consuming work to import and manipulate a very large file with R on personal computer. 
It'll be more efficiency to extract the target ranges from origin VCF with `filterLargeVCF()` before import. 
If your VCF file is just a few 'MB', this step was not necessary at all.

**Note:** if extract more than one ranges, length of output file names (`VCFout`) must be equal with `Chr` and `POS`.

```{r preprocess large VCF, eval=FALSE, include=TRUE}
# new VCF file will be saved to disk
# extract a single gene/range from a large vcf
filterLargeVCF(VCFin = "Ori.vcf.gz",
               VCFout = "filtered.vcf.gz",
               Chr = "scaffold_8",
               POS = c(19802,24501),
               override = TRUE)

# extract multi genes/ranges from large vcf
filterLargeVCF(VCFin = "Ori.vcf.gz",          # surfix should be .vcf.gz or .vcf
               VCFout = c("filtered1.vcf.gz", # surfix should be .vcf.gz or .vcf
                          "filtered2.vcf.gz", 
                          "filtered3.vcf.gz"),
               Chr = c("scaffold_8",
                       "scaffold_8",
                       "scaffold_7"),
               POS = list(c(19802,24501), 
                          c(27341,28949),
                          c(38469,40344)),
               override = TRUE)               # if TRUE, existed file will be override without warning
```

### DNA sequences manipulation

The origin DNA sequences must be aligned and trimmed due to haplotype calculation need all sequences have same length. Those operations could be done with `geneHapR`. I still suggest users align and trim DNA sequences with **Mega** software and then save the result as *FASTA* format before import them into R.

```{r DNA alignment and trim, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# sequences alignment
seqs <- allignSeqs(seqs, quiet = TRUE)

# sequences trim
seqs <- trimSeqs(seqs, minFlankFraction = 0.1)
seqs
```

## Haplotype calculation

As mentioned before, haplotype could be calculated from VCF or sequences with `vcf2hap()` or `seqs2hap()`. The genotype of most sites should be known and homozygous, still, a few site are unknown or heterozygous due to chromosome variant or error cased by sequencing or SNP calling or gaps or other reasons. It's a hard decision whether to drop accessions/individuals contains heterozygous or unknown sites for every haplotype analysis. Hence, I leave the choice to users.



Calculate haplotype result from VCF.
```{r haplotype calculation from vcf, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
hapResult <- vcf2hap(vcf,
                     hapPrefix = "H",
                     hyb_remove = TRUE,
                     na.drop = TRUE)
hapResult
```



Calculate haplotype result from aligned DNA sequences.

```{r haplotype calculation from seqs, eval=FALSE}
hapResult <- seqs2hap(seqs,
                      Ref = names(seqs)[1],
                      hapPrefix = "H",
                      hyb_remove = TRUE,
                      na.drop = TRUE,
                      maxGapsPerSeq = 0.25)
```

## Adjustment of `hapResult`

Before visualization, there were a few details need to be adjusted.
eg. add annotations and adjust position of "ATG"

### Add annotations to `hapResult`

While `hapResult` was calculated from `vcfR` object, the **INFO** was taken from `@fix` field. The *VCF INFO* may missing some annotations. or contents format was inappropriate to display. Further more, **INFO** contents nothing if `hapResult` was generated from sequences. Here, we can introduce/replace the origin **INFO** by `addINFO()`.

**Note** that: length of `values` must be equal with number of sites.

Let's see how mant sites contains in the `hapResult`.
```{r check site numbers}
# Chech number of sites conclude in hapResult
sites(hapResult)
```

Now we replace the old INFO field with new tag named as "PrChange".

```{r replace INFO, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
# add annotations to INFO field
hapResult <-addINFO(hapResult,
                     tag = "PrChange",
                     values = rep(c("C->D", "V->R", "G->N"),3),
                     replace = TRUE)
```

Here, we add a tag named as "CDSChange" followed the old INFO.

```{r add INFO, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
# To replace the origin INFO by set 'replace' as TRUE
hapResult <-addINFO(hapResult,
                     tag = "CDSChange",
                     values = rep(c("C->A", "T->C", "G->T"),3),
                     replace = FALSE)
```

### Adjust position of "ATG"

This function was only used to adjust the position of "ATG" to 0 and hence convert the gene on negative strand to positive strand.

Be note that: **GFF** and **hapResult** need to adjust position of *ATG* with the same parameters.

```{r ATG position, eval=FALSE}
# set ATG position as zero in gff
newgff <- gffSetATGas0(gff = gff, hap = hapResult,
                       geneID = "test1G0387",
                       Chr = "scaffold_1",
                       POS = c(4300, 7910)) 

# set position of ATG as zero in hapResult/hapSummary
newhap <- hapSetATGas0(gff = gff, hap = hapResult,
                       geneID = "test1G0387",
                       Chr = "scaffold_1",
                       POS = c(4300, 7910))
```

## `hapResult` summary and visualization

Once we have the `hapResult` object, can we summarize and visualize our `hapResult` by interact with annotations and phenotypes.

### Summary hapResult

Now, we have the `hapResult` object with INFOs we want display in next step.
The `hap_summary()` function convert the object of `hapResult` class, which is  a long table format, into a short table belong to `hapSummary` class. 
In `hapResult` each row represent a accession, while each row represents a hap in `hapSummary`.

```{r hapSummary, fig.height=4, fig.width=6, message=FALSE, warning=FALSE, paged.print=FALSE}
hapSummary <- hap_summary(hapResult)
hapSummary
```

### Visualize haplotye as table

Let's see how to visualization of our haplotype results.

At first let's display the `hapSummary` as a table.
In this table like figure we can see all the variants and their positions, haplotypes and their frequencies.

```{r, fig.height=4, fig.width=6, message=FALSE, warning=FALSE, paged.print=FALSE}
plotHapTable(hapSummary)
```

Also we can add an annotation, "CDSChange", to the table by assign the `INFO_tag`.
It's your responsibility to verify whether the INFO_tag was existed in the INFO field.

```{r, fig.height=4, fig.width=6, message=FALSE, warning=FALSE, paged.print=FALSE}
# add one annotation
plotHapTable(hapSummary,
             hapPrefix = "H",
             INFO_tag = "CDSChange", 
             tag_name = "CDS",
             displayIndelSize = 1, 
             angle = 45,
             replaceMultiAllele = TRUE,
             ALLELE.color = "grey90")
```

Now let's add another `INFO_tag` named as "PrChange".

```{r , fig.height=4, fig.width=6, message=FALSE, warning=FALSE, paged.print=FALSE}
# add multi annotation
plotHapTable(hapSummary,
             hapPrefix = "H",
             INFO_tag = c("CDSChange", "PrChange"),
             displayIndelSize = 1, 
             angle = 45,
             replaceMultiAllele = TRUE,
             ALLELE.color = "grey90")
```

Parameter `tag_name` was used to replace the character if `INFO_tag` was too long.
```{r, fig.height=4, fig.width=6, message=FALSE, warning=FALSE, paged.print=FALSE}
# add multi annotation
plotHapTable(hapSummary,
             hapPrefix = "H",
             INFO_tag = c("CDSChange", "PrChange"),
             tag_name = c("CDS", "Pr"),
             displayIndelSize = 1, 
             angle = 45,
             replaceMultiAllele = TRUE,
             ALLELE.color = "grey90")
```


### Display variations on gene model.

I think it's a good idea to figure out where are the variants by marking them on gene model.

```{r display variations on gene model, fig.height=4, fig.width=6, message=FALSE, warning=FALSE, paged.print=FALSE}
displayVarOnGeneModel(gff,
                      hapSummary,
                      Chr = "scaffold_1",
                      startPOS = 4300, endPOS = 7910,
                      type = "pin", cex = 0.7,
                      CDS_h = 0.05, fiveUTR_h = 0.02, threeUTR_h = 0.01)
```

### `hapNet` calculation and visualization

The `hapNet` could be generated from object of `hapSummary` class.
The accession group information could be attached in this step.

```{r hapNet}
hapNet <- get_hapNet(hapSummary,
                     accGroup = accGroup,
                     groupName = "Type")
```

Once we have the `hapNet` object, we can plot it with R.

```{r, fig.height=4, fig.width=6, message=FALSE, warning=FALSE, paged.print=FALSE}
# plot haploNet
plotHapNet(hapNet,
           size = "freq",                   # circle size
           scale = "log10",                 # scale circle with 'log10(size + 1)'
           cex = 0.8,                       # size of hap symbol
           col.link = 2,                    # link colors
           link.width = 2,                  # link widths
           show.mutation = 2,               # mutation types one of c(0,1,2,3)
           legendPosition = "right")        # legend position
```

### Phenotype association analysis

Finally, let's see which haplotype has superiority at particular area by interact with phynotype.

Here are two options, merged or separated, to organized the heatmap of p-values and violin plot.
The figure as an object of `ggplot2`, which means user could add/modified figure elements with `ggplot2`.

Here is an example for merged arrangement:

```{r hapVsPheno merged, fig.height=4, fig.width=10, message=FALSE, warning=FALSE, paged.print=FALSE}
results <-hapVsPheno(hapResult,
                     hapPrefix = "H",
                     title = "This is title",
                     mergeFigs = TRUE,
                     pheno = pheno,
                     phenoName = "GrainWeight.2021",
                     minAcc = 3)
plot(results$figs)
```



**An example for separated plot:**

```{r hapVsPheno separated, fig.height=4, fig.width=6, message=FALSE, warning=FALSE, paged.print=FALSE}
results <- hapVsPheno(hap = hapResult,
                      hapPrefix = "H",
                      title = "This is title",
                      pheno = pheno,
                      phenoName = "GrainWeight.2021",
                      minAcc = 3,
                      mergeFigs = FALSE)
plot(results$fig_pvalue)
plot(results$fig_Violin)
```

### Association analysis of multi-phenotypes once a time

I believe the function of `hapVsPhenos()` will be useful when you have a lot of phenotype need to be associated with haplotype results.

Note that: the pheno name will be added between the file name and surfix.  

```{r eval=FALSE}
hapVsPhenos(hapResult,
            pheno,
            outPutSingleFile = TRUE,
            hapPrefix = "H",
            title = "Seita.0G000000",
            file = "mypheno.tiff",
            width = 12,
            height = 8,
            res = 300)
```
